---
title: 'BIOL 720 Assignment #4'
author: "Elise Millar"
date: "November 17, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Course: Data Visualization with ggplot2 (Part 1)

# Chapter 1: Introduction

# 1) Intro

- Linear model not suitable for all data, depending on trend.
- If rely solely on numerical output w/o plotting data, miss these underlyig trends.

# 2) Explore + explain

- Exploratory plots are typically not pretty.

# 3) Exploring ggplot2, Part 1

```{r}
library(ggplot2)
str(mtcars)

ggplot(mtcars, aes(x = cyl, y = mpg) + geom_point()
       
```

- Gives a plot of cars binned into 4,6,8 cylinders but nothing for 5 +7 (don't exist).
- cyl (# cylinders) is categorical, but classified as numeric in mtcars (# vs. name).

# 4) Exploring ggplot2, Part 2

- Change previous to:

```{r}

ggplot(mtcars, aes(x = factor(cyl), y = mpg) + geom_point()
       
```
- Now binned into only 4,6,8.

# 5) Grammar of Graphics

2 Principles:
- Graphics = distinct layers of grammatical elements.
- Meaningful plots through aesthetic mapping.

Essential Elements:
- Data = dataset being plotted.
- Aesthetics = scales onto which we map data.
- Geometries = visual elements used for data.

# 6) Exploring ggplot2, Part 3

- To make colour dependent on displacement of car engine (disp).

```{r}

ggplot(mtcars, aes(x = wt, y = mpg, col = disp)) + geom_point()
       
```

# 7) Understanding Variables

- A continuous variable, such as disp, cannot be mapped to shape.
- Shape has only a finite # of shapes (needs catgorical data).

# 8) ggpolt2

1st layer: Data - species, sepal length, sepal width, petal length, petal width.
2nd layer: Aesthetics - x = sepal length, y = sepal width.
3rd layer: Geometries - how the plot will look.

- If the first plot is messy, need to use the other layers.

4th layer: Facets - want to make 3 separate plots, one for each species (setosa, versicolor, virginica)
5th layer: Statistics - add parameters, like a linear model to subplots.
6th layer: Coordinates - specify dimensions of plot + clean up labelling + scaling of axes.
7th layer: Theme - all non-data ink, making a publishable plot in R.

# 9) Exploring ggplot2, Part 4

geom_point() - draws points on the plot.
geom_smooth() - draws smoothed line over points.

# 10) Exploring ggplot2, Part 5

- Uisng only geom_smooth = only smooth line no points.
- Can colour according to clarity of diamonds by making color = clarity.

Can make translucent coloured points by:
- removing smooth layer (geom_smooth())
- adding points layer (geom_point())
- set alpha = 0.4 inside geom_point() (makes points 60% transparent + 40% visible)

Will get a very pretty plot!
 
# 11) Understanding grammar, Part 1

- Can put basic info in an object.
- And put attributes into geom layer

```{r}

dia_plot <- ggplot(diamonds, aes(x = carat, y = price))

dia_plot + geom_point()

dia_plot + geom_point(aes(col = clarity))
       
```

# 12) Understanding grammar, Part 2

- Can add geom_smooth() layer on top.
- Exclude error shading using se = FALSE.
- Can modify to contain aes(), map col to clarity.

```{r}

dia_plot <- dia_plot + geom_point(alpha = 0.2)

dia_plot + geom_smooth(se = FALSE)

dia_plot + geom_smooth(aes(col = clarity), se = FALSE)
       
```


# Chapter 2: Data

# 1) Objects + layers

Base plot vs. ggplot2

Base plot:
- plot sepal length vs width
- if want to ass petal length + width, need to add as points + colour them.
   - plot doesn't get redrawn
   - draws plot as image
   - need to manually add legend
- can only make limited kinds of plots.

ggplot2:
- used for all types of plots or even custom plots.
- if assign base layers to object, can recycle object w variety of diff plot types.

# 2) Base package + ggplot2, Part 1 - Plot

- Can add new column to mtcars, fcyl. 

```{r}

plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)

mtcars$fcyl <- as.factor(mtcars$cyl)

plot(mtcars$wt, mtcars$mpg, col = mtcars$fcyl)
       
```

- Changed from cyl = 4,6,8 to fcyl = 1,2,3.

# 3) Base package + ggplot2, Part 2 - lm

lm() = linear model
lapply() = input is a vector + function. Applies the function to each element of the vector + returns a list.

Ex. Can vectorize lm() over cyl variable using lapply().
- lapply() takes each element of mtcars$cyl + calls the function in the arguement.
- takes a value of mtcars$cyl + subsets the data so only rows w cyl == x are used.
- fits linear model to filtered datset + uses that model to add a line to the plot w the abline() function.
- can also make legend.

```{r}

carModel <- lm(mpg ~ wt, data = mtcars)

abline(carModel, lty = 2)

plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)
lapply(mtcars$cyl, function(x) {
  abline(lm(mpg ~ wt, mtcars, subset = (cyl == x)), col = x)
})

legend(x = 5, y = 33, legend = levels(mtcars$cyl), col = 1:3, pch = 1, bty = "n")
       
```

# 4) Base package + ggplot2, Part 3

- Including the lines of linear models, per cyl.
- Including a lm for the entire dataset in its whole.

```{r}

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point()

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, linetype = 2)
       
```

# 5) ggplot2 compared to base package

ggplot2:
- creates plotting objects, which can be manipulated.
- takes care of a lot of leg work (legends, colours).
- built upon grammar of graphics plotting, allows better understanding b/w data + visual relationships.

# 6) Proper Data Format

ggplot2:
- plot sepal length vs width.
- add in peta length + width via geom_point.
  - plotting space is adjusted
  - produces a manipulatable object

Colour = scale
-  normally:

```{r}

ggplot(iris.wide, aes(x = Length, y = Width, col = Part)) +
  geom_point()
       
```

iris = both plant part + measuremnt split amongst groups.
iris.wide = measurement split across columns.

# 7) Plotting the ggplot2 way

Using the method of ggplot2 vs Base R is better for this data. Use iris.wide.

# 8) Tidy Data

iris.wide2 = species split amongst columns.

Can do either:

```{r}

ggplot(iris.wide2, aes(x = Part, y = Setosa, col = Measure)) +
  geom_jitter()

# OR

ggplot(iris.wide2, aes(x = Measure, y = Setosa, col = Part)) +
  geom_jitter()
       
```

If want to do all 3 species (not just Setosa, all at once) need tody data:
- each row contains single observation.
- each column contains a single variable.
  - tody format (access to all variables in dataset)

facet_grid() - can compare all species in one plot (small mutiples)
             - all on same axes (yay!)
             
*Depends on how mnay axes you have

# 9) Variables to visuals, Part 1a

iris vs. iris.wide vs. iris.tidy
- iris.tidy has least amount of variables + most obs, most accessible. 

```{r}

ggplot(iris.tidy, aes(x = Species, y = Value, col = Part)) +
  geom_jitter() +
  facet_grid(. ~ Measure)
       
```

# 10) Variables to visuals, Part 1b

gather() - rearranges data frame by specifying columns that are catergoical variables with a -
separate() - splits up new key column, containing former headers, according to ..

Producing tidy data:

```{r}

iris.tidy <- iris %>%
  gather(key, Value, -Species) %>%
  separate(key, c("Part", "Measure""), "\\.")
       
```

# 11) Variables to visuals, Part 2a

Most suitable iris.wide plot:

```{r}

ggplot(iris.wide, aes(x = Length, y = Width, col = Part)) +
  geom_jitter() +
  facet_grid(. ~ Species)
       
```

# 12) Variables to visuals, Part 2b

Producing iris.tidy:

- Add column w unique idea
- Flower + Species are categorical.
- Column names are Measure + Part.
- spread() distributes new Measure column + value column into 2 columns.

```{r}

iris.$Flower <- 1:now(iris)

iris.wide <- iris %>%
  gather(key, value, -Species, -Flower) %>% 
  separate(key, c("Part", "Measure"), "\\.") %>%
  
  spread(Measure, value)
       
```


# Chapter 3: Aesthetics

# 1) Visible Aesthetics 

- Normally aesthetics = colour, shape, size (how it looks)
- Here it is an attribute:
  - use asethetics for which variable is mapped onto it (Ex. Species mapped onto colour)
- Use aesthetics to maximize variables on a  plot.
- x + y are aesthetics:
  - Ex. sepal length mapped on x-axis + width mapped on y-axis
- Changing colour of dots using col changes a visible attribute not the aesthetics.
- Can add a 3rd variable, map species onto colour (visible aesthetic).
- Aesthetics called in aes() + attributes called in geom_():
  - can call aes() in geom layer
  - only done if need to compare diff data sources
  
Typical aesthetics:
x = x axis position
y = y axis position
colour = colour of dots, outlines of other shapes
fill = fill colour
size = diameter of points, thickness of lines
alpha = trasnparency (0-1)
linetype = line dash pattern
labels = text on a plot or axes
shape = shape of a point

# 2) All about aesthetics, Part 1

```{r}

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point(shape = 1, size = 4)
       
```

# 3) All about aesthetics, Part 2

shape = 19 - solid circle w outline same colour as inside.
shape = 1 - hollow circle (no fill)
shape = 16 - solid, no outline

*Remember to set alpha for solid points = 1

shape = 21 - use both fill for inside + col for outline. Good for mapping 2 aesthetics to a dot.

```{r}

ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl, col = am)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)
       
```

- Use fill when have another shape (a bar) or when using a point that has a fill + colour attribute (shape = 21 - circle w an outline).
- New plot with col=am has 2 attributes for single dot.

# 4) All about aesthetics, Part 3

```{r}

ggplot(mtcars, aes(x = wt, y = mpg, size = cyl)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)

# vs. shape = cyl, alpha = cyl

ggplot(mtcars, aes(x = wt, y = mpg, label = cyl)) +
  geom_text()

#shows names of cylinders (4,6,8)
       
```

# 5) All about attributes, Part 1

shapes = 1-20 - can only accept a colour aesthetic
shapes = 21-25 - have both a color + fill aesthetic

```{r}

my_color <- ."#4ABEFF"

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point()

#vs.

ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point(col = my_color)

# colour is overwritten

ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(col = my_color, size = 10, shape = 23)

# fill colours the inside of the dot, color outlines the dot
       
```

# 6) All about attributes, Part 2

```{r}

ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
  geom_point(alpha = 0.5)


ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl, col = am)) +
  geom_point(shape = 21, col = "yellow")


ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl, col = am)) +
  geom_text(label = rownames(mtcars), col = "red", size = 24)      
       
```

# 7) Going all out

- Variables in a data frame mapped to aesthetics in aes().
- Visual elements are set by attributes in specific geom layers (geom_point(col = "red"))

*Aesthetics

```{r}

ggplot(mtcars, aes(x = mpg, y = qsec, col = factor(cyl))) +
  geom_point()

ggplot(mtcars, aes(x = mpg, y = qsec, col = factor(cyl), shape = am)) +
  geom_point()

ggplot(mtcars, aes(x = mpg, y = qsec, col = factor(cyl), shape = am, size = (hp/wt))) +
  geom_point()   
       
```

# 8) Aesthetics for categorical + continous variables

- Label + shape are only categorical variables.

# 9) Modifying Aesthetics

Positions:
- identity
- jitter

Scale Functions:
- scale_x
- scale_y
- scale_color
- scale_fill
- scale_linetype
- scale_shape

Ex. scale_x_continuous, scale_color_discrete

- Name of scale, followed by limits, breaks, labels, expand

# 10) Position

```{r}

cyl.am <- ggplot(mtcars, aes(factor(cyl), fill = factor(am))) 

+ geom_bar() 
# Gives fat bars, differing heights, made of 2 colours

+ geom_bar(position = "fill")
# Gives bars reaching to top (proportions), made of 2 colours

+ geom_bar(position = "dodge")
# Gives bars of differing heights, colour split into their own bars
       
```

# 11) Setting a dummy aesthetic

Univariate data = fake y axis (no real y axis)

Error if only use aes(x = ..) and no y.

```{r}

ggplot(mtcars, aes(x = mpg, y = 0)) +
  geom_jitter()

ggplot(mtcars, aes(x = mpg, y = 0)) +
  geom_jitter() +
  scale_y_continuous(limits = c(-2, 2))
       
```

# .....took way too many notes, will be skipping over a few sub-chapters..


# Chapter 4: Geometries

# 2) Scatter plots + jittering

```{r}

ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter()

ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter(width = 0.1)

ggplot(mtcars, aes(x = cyl, y = wt)) +
  geom_jitter(position = position_jitter(0.1))
       
```

# 4) Bar plots

Histogram = shows bin distribution of cotinous variable.
  - only need one aesthetic, x
  - bins + not original data
  - plot of statistical function
- Can change binwidth
  - Make it larger, easier to see
  - geom_histogram(binwidth = 0.1)
- No space b/w bars: represnts continous distribution.

# 6) Position

Stack = place bars on top of each other. Counts used. Default.
Fill = place bars on top of each other, but use proportions.
Dodge = place bars next to each other, Counts used.

# 8) Overlapping histograms, Part 1

Frequency polygon = specific to binned date that draws a line connecting the value of each bin.
- takes binwidth arguement + by default stat = "bin" + position = "identity"

# 15) Multiple time series, Part 1

```{r}

str(fish.species)
#year + all species of salmon listed, too many variables

fish.tidy <- gather(fish.species, Species, Capture, -Year)
# now we will ahve 3 variables
       
```

# 16) Multiple time series, Part 2

```{r}

ggplot(fish.tidy, aes(x = Year, y = Capture, col = Species)) +
  geom_line()
       
```

# Also did Chapter 5: qplot + wrap-up. Notes are in notebook. Can refer to them later if ever using qplot.











